Towards Contract Modules for the Tezos Blockchain

### 2021-05-26 papers/contract-modules/fmbc2021/fmbc-main.tex
as submitted to fmbc

The contract works as follows. It is deployed with storage
\lstinline/(true,(owner,owner))/ which indicates that bidding is
allowed and the contract owner is currently also the highest
bidder. On deployment the owner also deposits an initial balance to indicate the
minimum bid. Closing the contract transfers the balance to the
owner. It is restricted to the owner. Bidding fails if the contract is
closed. If bidding is open and the amount of tokens with the bid exceeds the
current highest bid, the current bidder replaces the previous highest
bidder and the previous highest bidder is reimbursed. Otherwise,
bidding fails, too.

@@

Besides the obvious, technology induced ways that a contract invocation might fail
(insufficient gas price offered, insufficient gas to complete, timeout
due to lack of connectivity, etc) a Michelson contract can fail due to
a programmer induced condition. To this end, there is a Michelson
instruction \lstinline/FAILWITH/ that terminates contract execution
with an error message which is reported back to the caller. This error
message includes a rendering of the top value on the stack.

We consider the technological failures like Java's unchecked
exceptions, but we wish to deal with the explicit failures like
checked exceptions.
Our generated code handles failures in a suitable
error monad that makes the failures explicit in a custom
datatype.\footnote{Alternatively, this could be done using OCaml
  exceptions, but we chose to stay within the monadic framework that
  is already used by existing Tezos APIs.}

\begin{lstlisting}[float,captionpos=b,caption={Contract module example},label={lst:contract-module-example}]
contract type Auction = sig
  paid entrypoint bid () 
  raises "closed"               (** auction closed *)
       | "too low"              (** bid too low    *)

  entrypoint close ()
  raises "not owner"            (** caller cannot close *)
end
\end{lstlisting}
Listing~\ref{lst:contract-module-example} shows a contract module for
the auction contract.  It declares the entrypoint \lstinline/bid/ as
\lstinline/paid/, which means that it needs to be invoked with a
non-zero amount of tokens, it gives the pattern \lstinline/()/ for the
input value of type \lstinline/unit/, and it specifies two possible
failure messages that we wish to deal with programmatically.
The \lstinline/close/ entrypoint is similar, but requires no tokens.

\begin{lstlisting}[float,captionpos=b,caption={Generated signature},label={lst:generated-signature}]
module type MyContract = sig
  type bid_errors = 
     | closed       (** auction closed*)
     | too_low      (** bid received is too low *)

  val bid
    : Tezos.pukh -> ?fee:Tezos.tz -> amount:Tezos.tz
    -> (Tezos.status, bid_errors) Tezos.monad

  type close_errors = 
     | not_owner    (** caller cannot close the auction *)

  val close
    : Tezos.pukh -> ?fee:Tezos.tz
    -> (Tezos.status, close_errors) Tezos.monad
end
\end{lstlisting}
Listing~\ref{lst:generated-signature} contains an OCaml module
signature as it could be generated from the contract module. The
module \lstinline/Tezos/ supposedly contains types and other low-level
Tezos-specific definitions. Specifically, the type \lstinline/pukh/ for public
key hashes identifies contracts, the type \lstinline/tz/ stands for
Tezos tokens, the type \lstinline/status/ reflects the internal return
status, and \lstinline/monad/ is an internal monad type. The signature
declares a function and an error type for each entrypoint.

@@

\section{Advanced Checking}
\label{sec:advanced-checking}
\begin{lstlisting}[float,captionpos=b,caption={Safer contract module},label={lst:safer-contract-module}]
contract type SaferAuction = sig
  storage (Pair (bidding : bool) 
                (Pair (owner : address) (hi_bidder : address)))

  entrypoint close ()
  requires (sender = owner) raises "not owner"

  paid entrypoint bid () 
  requires bidding raises "closed" (** auction closed*)
  requires (amount > balance)
  raises "too low"                 (** bid too low *)
end
\end{lstlisting}

As it is expensive to invoke a contract just to find out that it
fails, we propose to extend entrypoint specifications with
preconditions as shown in Listing~\ref{lst:safer-contract-module}. The
idea is that the generated OCaml module tries to check the
preconditions off-chain before invoking the contract. To this end, the
code needs to obtain properties like balance, storage etc of the
contract, but this information is available from the Tezos node without a fee!
We discuss two of the preconditions to highlight the properties that 
need to be analyzed.

@@

The basic idea of our advanced checking, namely precondition checking,
is inspired by JML, the Java modeling
language~\cite{Leavens2006DesignBC, 101007}, in which the behavior of 
program components is described as a contract between Java
program and its clients. This contract specifies
preconditions that must be satisfied by clients and postconditions
that are guaranteed by the program. A precondition supplied with a
client call must be verified before a function defined by the program
is called, and the program guarantees that the postconditions are
satisfied in return after the call. The original idea of using
preconditions and postconditions dates back to Hoare's paper
\cite{101145}. Software contracts have also been proposed for
blockchain \cite{DBLP:journals/fbloc/Bartoletti20}.
% The executable contracts are written as program code in
% the programming language.
In our approach, the safe contract module in the Ocaml language comes
close to contracts in this sense. Several
applications are based on JML \cite{Tran2017}. Ahrendt et
al. \cite{keY} propose the KeY framework for deductive software
verification.

@@

Current blockchains often provide low-level interfaces to interact
with smart contracts. These interfaces work with loosely structured
without static guarantees. This paper presents ongoing research on the
programmatic interaction with smart contracts on the Tezos blockchain
that could benefit developers of mixed applications and
oracles comprised of on-chain and off-chain parts. The approach does
not provide a general API, but targets each 
individual smart contract by generating a specialized contract module
that provides a typed high-level interface from a contract
specification. In doing so, errors from contract calls are explicitly
specified in a user-defined data type. A contract call is wrapped in a
fully typed and integrated OCaml function. In addition, the wrapper can check preconditions before the 
actual call to reduce the waste of gas of a failed call.
% As future work, we
% combine the contract module with our typed API designed to provide a
% general type-safe programmatic API for the Tezos blockchain.

>>>>

The contract works as follows. It is deployed with storage
\lstinline/(true,(owner,owner))/ which indicates that bidding is
allowed and the contract owner is currently also the highest
bidder. On deployment the owner deposits an initial balance to indicate the
minimum bid. Closing the contract transfers the balance to the
owner. It is restricted to the owner. Closing as well as bidding fails
if the auction is
closed. If bidding is open and the amount of tokens accompanying the bid exceeds the
current highest bid, the current bidder replaces the previous highest
bidder and the previous highest bidder is reimbursed. Otherwise,
bidding fails, too.

@@

Besides the obvious, technology induced ways that a contract invocation might fail
(insufficient gas price offered, insufficient gas to complete, timeout
due to lack of connectivity, etc) a Michelson contract can fail due to
a programmer induced condition caused by the
instruction \lstinline/FAILWITH/. It terminates contract execution
with an error message which is reported back to the caller. This error
message includes the top value on the stack.

We consider the technological failures like Java's unchecked
exceptions, but we wish to deal with the explicit failures like
checked exceptions~\cite{DBLP:conf/oopsla/AnconaLZ01}.
Our generated code handles failures in a suitable
error monad that makes the failures explicit in a custom
datatype.\footnote{Alternatively, this could be done using OCaml
  exceptions, but we chose to stay within the monadic framework that
  is already used by existing Tezos APIs.}

\begin{lstlisting}[float,captionpos=b,caption={Example contract module},label={lst:contract-module-example}]
contract type Auction = sig
  paid entrypoint bid () 
  raises "closed"               (** auction closed *)
       | "too low"              (** bid too low    *)

  entrypoint close ()
  raises "closed"               (** auction closed *)
       | "not owner"            (** caller cannot close *)
end
\end{lstlisting}
Listing~\ref{lst:contract-module-example} shows a contract module for
the auction contract.  It declares the entrypoint \lstinline/bid/ as
\lstinline/paid/, i.e., it needs to be invoked with a
non-zero amount of tokens, it gives the pattern \lstinline/()/ for the
input value of type \lstinline/unit/, and it specifies two possible
failure messages that we wish to deal with programmatically.
The \lstinline/close/ entrypoint is similar, but requires no tokens.

% module type MyContract = sig 
\begin{lstlisting}[float,captionpos=b,caption={Generated signature},label={lst:generated-signature}]
  type bid_errors = 
     | bid_closed      (** auction closed*)
     | bid_too_low     (** bid received is too low *)

  val bid
    : Tezos.pukh -> ?fee:Tezos.mutez -> amount:Tezos.mutez
    -> (Tezos.status, bid_errors) Tezos.monad

  type close_errors = 
     | close_closed    (** auction closed *)
     | close_not_owner (** caller cannot close the auction *)

  val close
    : Tezos.pukh -> ?fee:Tezos.mutez
    -> (Tezos.status, close_errors) Tezos.monad
\end{lstlisting}
% end 
Listing~\ref{lst:generated-signature} contains an OCaml module
signature as it could be generated from the contract module. The
module \lstinline/Tezos/ supposedly contains types and other low-level
Tezos-specific definitions. The type \lstinline/pukh/ for public
key hashes identifies contracts, the type \lstinline/mutez/ stands for
Tezos tokens, the type \lstinline/status/ reflects the internal return
status, and \lstinline/monad/ is an internal monad type. The signature
declares a function and an error type for each entrypoint.

@@

\section{Advanced Checking}
\label{sec:advanced-checking}
\begin{lstlisting}[float,captionpos=b,caption={Enhanced contract module},label={lst:safer-contract-module}]
contract type SaferAuction = sig
  storage (Pair (bidding : bool) 
                (Pair (owner : address) (hi_bidder : address)))

  entrypoint close ()
  requires bidding raises "closed" (** auction closed*)
  ensures not bidding
  requires (sender = owner) raises "not owner"

  paid entrypoint bid () 
  requires bidding raises "closed" (** auction closed*)
  ensures bidding
  requires (amount > old.balance)
  raises "too low"                 (** bid too low *)
  ensures (balance >= old.balance)
  ensures (owner = old.owner)
end
\end{lstlisting}

As it is expensive to invoke a contract just to find out that it
fails, we propose to extend entrypoint specifications with
preconditions as shown in Listing~\ref{lst:safer-contract-module}. The
idea is that the generated OCaml module tries to check the
preconditions off-chain before invoking the contract. To this end, the
off-chain code needs to obtain properties like balance, storage etc of the
contract, but this information is available from the Tezos node without a fee!
We discuss two of the preconditions to highlight the properties that 
need to be analyzed.

@@

The basic idea of our advanced checking, namely precondition checking,
is inspired by JML, the Java modeling
language~\cite{Leavens2006DesignBC, 101007}, in which the behavior of 
program components is described as a contract between Java
program and its clients. This contract specifies
preconditions that must be satisfied by clients and postconditions
that are guaranteed by the program. A precondition supplied with a
client call must be verified before a function defined by the program
is called, and the program guarantees that the postconditions are
satisfied in return after the call. The original idea of using
preconditions and postconditions dates back to Hoare's paper
\cite{101145}. Software contracts have also been proposed for
blockchain \cite{DBLP:journals/fbloc/Bartoletti20}.
% The executable contracts are written as program code in
% the programming language.
In our approach, the safe contract module in the OCaml language comes
close to contracts in this sense. Several
applications are based on JML \cite{Tran2017}. Ahrendt et
al. \cite{keY} propose the KeY framework for deductive software
verification.

@@

Current blockchains often provide low-level interfaces to interact
with smart contracts. These interfaces work with loosely structured
without static guarantees. This paper presents ongoing research on the
programmatic interaction with smart contracts on the Tezos blockchain
that could benefit developers of mixed applications and
oracles comprised of on-chain and off-chain parts. The approach does
not provide a general API, but targets each 
individual smart contract by generating a specialized contract module
that provides a typed high-level interface from a contract
specification. In doing so, errors from contract calls are explicitly
specified in a user-defined data type. A contract call is wrapped in a
fully typed and integrated OCaml function. In addition, the wrapper can check preconditions before the 
actual call to reduce the waste of gas of a failed call.

Status: we are currently working on a prototype of the symbolic interpreter.
% As future work, we
% combine the contract module with our typed API designed to provide a
% general type-safe programmatic API for the Tezos blockchain.

### 2021-05-26 papers/contract-modules/fmbc2021/macros.tex
as submitted to fmbc

\lstset{
  language=Caml,
  basicstyle=\ttfamily,
  backgroundcolor=\color{pypypy},
  morekeywords={module,contract,sig,val,paid,entrypoint,raises},
  classoffset=1,
  morekeywords={nat,int,unit,string,or,pair,option,address,bool},
  morekeywords={Tezos,pukh,tz,status,monad},
  keywordstyle=\color{blue},
}

>>>>

\lstset{
  language=Caml,
  basicstyle=\ttfamily,
  backgroundcolor=\color{pypypy},
  morekeywords={requires,ensures,sender,amount,balance,storage,hash,module,contract,sig,val,paid,entrypoint,raises},
  classoffset=1,
  morekeywords={nat,int,unit,string,or,pair,option,address,bool},
  morekeywords={Tezos,pukh,mutez,status,monad},
  keywordstyle=\color{blue},
}
